Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-04-17 pituba/article/article.tex
more and more

Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume. The integration process, however, requires all quantities to be either spatially averaged or assumed constant.

@@

We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).

All quantities are locally functions of space and time, but, after averaged in space, they become functions of time only:

@@

Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} rewritten in terms of averages read:

@@

We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of the same phase at standard pressure and temperature, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard condition:

@@

@@

Material balance equation is widely employed to stablish a relation between production (and injection) and reservoir pressure. It turns out, as we know, that pressure is different for different locations, mostly for different values of depth, and also for different phases.

@@

We shall see that for the purpose of computing the averaged value of phase densities in material balance, we may define some reference depth and use the corresponding local pressure without introducing any important error. Phase pressure increases with depth according to the hydrostatic gradient. The pressure increase, however, is not significant to produce appreciable variation in density.

@@

We may quantify this approximation by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

Integrating above equations between times $t_1$ and $t_2$ gives:
% Water
\begin{equation}\label{eq: Sw5}
\frac{V_{p2} \bar{S}_{w2}}{B_{w2}}=W+\left(W_i+W_e-W_p\right) \, ,
\end{equation}

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except at bubble-point pressure.
The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase. Each phase parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are clearly non-negative values. Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).

>>>>

Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume. The integration process requires all quantities to be either spatially averaged or assumed constant.

@@

We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.

All quantities are locally functions of space and time that, after averaged in space, become functions of time only:

@@

Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} rewritten in terms of averaged quantities read:

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of the same phase at standard pressure and temperature, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in standard condition:

@@

Integrating Eqs.~\eqref{eq: Sw4}, \eqref{eq: So4} and \eqref{eq: Sg4} between times $t_1$ and $t_2$ gives:
% Water
\begin{equation}\label{eq: Sw5}
\frac{V_{p2} \bar{S}_{w2}}{B_{w2}}=W+\left(W_i+W_e-W_p\right) \, ,
\end{equation}

@@

Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure. It turns out, as we know, that pressure is different for different locations, mostly for different depths, and also for different phases.

@@

We shall see that for the purpose of computing the averaged value of phase density in material balance, we may define some reference depth and use the corresponding local pressure as an approximation to the averaged density without introducing any important error. Although phase pressure increases with depth according to the hydrostatic gradient, this increase is not significant to produce appreciable variation in density.

@@

This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase. Each phase parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are non-negative values. Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).
